---
title: 'Modelling Daily US COVID deaths'
subtitle: 'Discussion Week 11 DATA621'
author: 'Bonnie Cooper'
output:
  rmdformats::downcute
---

Attempting autoregression to model the US Daily COVID Deaths.  
Will demonstrate an Ordinary Least Squares (OLS) approach and a Generalized Least Squares (GLS) models.  
How can I improve my approach here? ....


## Importing the data
start with taking a look at the dataset `us_covid19_daily.csv` which is freely available for download from [kaggle](https://www.kaggle.com/sudalairajkumar/covid19-in-usa?select=us_covid19_daily.csv).
```{r message=FALSE}
library( dplyr )
library( ggplot2 )
library( nlme )
df <- read.csv( '/home/bonzilla/Documents/MSDS/DATA621/us_covid19_daily.csv')
df <- transform(df, date = as.Date(as.character(date), "%Y%m%d"))
glimpse( df )
```

When we visualize the daily increase in deaths, the weekly cyclical behavior of the data is obvious:
```{r fig.cap="Daily COVID deaths in the US"}
COVID_ddp <- ggplot( df, aes( x = date, y = deathIncrease ) ) +
  geom_line() +
  theme_classic() +
  ggtitle( 'Daily US COVID deaths' ) +
  ylab( 'daily COVID deaths' )
COVID_ddp
```

To try to capture this behavior I apply the same autoregression method from LMR section 4.3.
```{r}
lagdeaths <- embed( df$deathIncrease, 9 )
colnames( lagdeaths ) <- c( 'death', paste0('lag',1:8) )
lagdeaths <- data.frame( lagdeaths )
glimpse( lagdeaths )
```

## Autoreggressive Model Fit

Fit a linear model that incorporates the lagged values as an autoregressive process
```{r}
armod <- lm( death ~ lag1 + lag7 + lag8, lagdeaths )
summary( armod )
```

All three of the lagged variables are strongly significant and the $R^2$ value suggests that the model accounts for for quite a bit of the variance in the data.  

Now to visualize the model fit with the data:
```{r fig.cap="Autoregression of Daily Death that considers several past time points. OLS = red, data = black"}
death_lag <- df[9:320,]
death_lag[ 'predict' ] <- predict( armod ) 
COVID_ddp_ar <- COVID_ddp +
  geom_line( aes( x = date, y = predict, color = 'red' ), data = death_lag, show.legend = FALSE )
COVID_ddp_ar
```

The the autoregressive linear model prediction follows the envelope of the data reasonably well.

## OLS Model Fit
```{r}
armod_ols <- lm( deathIncrease ~ hospitalizedCurrently + 
                    inIcuCurrently + onVentilatorCurrently +
                    hospitalizedIncrease + negativeIncrease + positiveIncrease +
                    totalTestResultsIncrease + positive + negative + recovered + 
                   death + hospitalized + totalTestResults, 
                  data = df )
summary( armod_ols )
```
```{r fig.cap="Standard OLS linear regression fit for daily US COVID deaths. OLS = blue, autoreg = red, data = black"}
df2 <- df[-armod_ols$na.action, ]
death_lag2 <- df2
death_lag2[ 'predict_ols' ] <- predict( armod_ols ) 
COVID_ddp_ols <- COVID_ddp +
  geom_line( aes( x = date, y = predict ), data = death_lag, show.legend = FALSE, color = 'red', size = 0.2 ) +
  geom_line( aes( x = date, y = predict_ols ), data = death_lag2, show.legend = FALSE, color = "blue", size = 1 ) 
COVID_ddp_ols
```

## GLS Model Fit

```{r}
armod_gls <- gls( deathIncrease ~ hospitalizedCurrently + 
                    inIcuCurrently + onVentilatorCurrently +
                    hospitalizedIncrease + negativeIncrease + positiveIncrease +
                    totalTestResultsIncrease + positive + negative + recovered + 
                    death + hospitalized + totalTestResults, 
                  correlation = corAR1( form = ~ 1 ),
                  data = na.omit( df ) )
summary( armod_gls )
```
```{r fig.cap="GLS fit for daily US COVID deaths. GLS = green, OLS = blue, autoreg = red, data = black"}
death_lag3 <- na.omit( df )
death_lag3[ 'predict_gls' ] <- predict( armod_gls ) 
COVID_ddp_gls <- COVID_ddp +
  geom_line( aes( x = date, y = predict ), data = death_lag, show.legend = FALSE, color = 'red', size = 0.2 ) +
  geom_line( aes( x = date, y = predict_ols ), data = death_lag2, show.legend = FALSE, color = "blue", size = 0.2 ) +  
  geom_line( aes( x = date, y = predict_gls ), data = death_lag3, show.legend = FALSE, color = 'green', size = 1 ) 
COVID_ddp_gls
```

...and that is where I've left off  

What are some next steps?  

1. combine the lagged features from the 1st autoregression with the original data for another OLS approach
2. am I using the `gls()` function correctly?...will RTFM to see if I can make improvements with that approach
3. ....any siggestions?....
4. comparing models





<br><br><br>
